Estimating duration in the presence of misclassification

### 2024-01-17 cis-imperfect-testing.tex
proof-read

Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} and the observed CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$, the beginning and end day of the episode both being equal to the day of this one positive test.
If this is true, then the duration $d_i = e_i - b_i + 1 = 1$.
Additionally, short episodes are very likely to be missed; therefore, the data is compatible with a large number of missed short episodes.
Therefore, single positive episodes suggest 
I confirmed this hypothesis by estimating the survival function in one of these simulated datasets with the addition of even a few single positives.
This produced the same pattern, that is estimating too many short episodes (not shown).

\Textcite{shenNonparametrica}, extending \textcite{panNote}, studied the situation when only the terminating event is interval censored and the initiating event has simple left truncation.
This is a related, although simpler, setting than the CIS.
The CIS has double interval censoring and a more complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
\Textcite{shenNonparametrica} showed that, in their situation, the maximum likelihood estimator is inconsistent if there is at least one single positive episode (check the precise statement) and can lead to a severely biased underestimation of the survival function.
While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are two strands of evidence supporting this hypothesis.
First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully, and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
The simulation setup expands the method of \cref{E-perf-test:sec:simulation-study}, including false negatives (see \cref{imperf-test:sec:simulate} for more details).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this section, I introduce a probabilistic model for the generation of false negatives.
The simplest model for false negatives is a constant probability of returning a negative test even when detectable.
The probability of testing positive given being detectable is known as the \emph{test sensitivity}, $\psens$.
$\psens = 1$ means that there are no false negatives, the situation previously considered.
\todo{check if detectable and test sensitivity defined somewhere, and in a consistent way}
In \cref{E-perf-test}, the outcome of a test was deterministic, positive if the individual being tested had an ongoing infection episode.
Now, the outcome of such a test result is a Bernoulli random variable with a probability $\psens$ of producing a positive.
If $\psens = 1$, this recovers the previous model without false negatives.
As previously, tests in individuals without ongoing infection episodes are always negative.

I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to include the possibility of false negatives.
More single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 93--96\%).
The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
The requirement for a much lower value of $\psens$ in the simulation study suggests this model of false negatives is inadequate.

More plausibly, the rate of false negatives varies over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

The CIS data provides evidence for a test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by classifying all test results as follows.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
This generates broad bounds, although suggest a declining test sensitivity over the course of an episode.
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

To check sensitivity of the inference, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
  \label{imperf-test:eq:variable-test-sensitivity}
\end{equation}
where $t$ is the number of days since infection.
Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.

@@

First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode being in the admissible, to allow for the episode possibly being longer than observed.
Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode being undetected, to allow for additional episodes being missed.

@@

I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
If it is a false negative, then I will consider the episode's end right-censored.
However, if it is a true negative, then the episode's end is interval-censored as previously.
A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.

I start by finding the tests that need consideration.
For tractability, assume that the negative test bounding the start of the infection, on day $l_i^{(b)}-1$, is a true negative.
This assumption is reasonable because the test sensitivity is high early in an infection; therefore, this test is unlikely to be a false negative.
For tractability and simplicity, I consider only tests between the positive tests providing a lower bound on the length of the episode.
By definition (see\todo{insert where these are defined in chapter 2}), these are the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive.
They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
Denote this subset as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq l_i^{(e)} \}$, and their results by $\vec{y_i'} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
As I assume that there are no false positives, the infection episode must span at least this period starting and ending with a positive test.
Therefore, the test results at times in $\sched_i'$ are either true positives or false negatives.

@@

First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
These occur with probability 1, as I continue to assume no false positives.
The tests in $\sched_i'$ are either true positives or false negatives.
\begin{align}
&p(\vec{y_i'}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(\vec{y_i'}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(\vec{y_i'} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
\end{align}

@@

False negatives add additional mechanisms by which episode $i$ could be undetected, in addition to those considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
Specifically, episode $i$ could be undetected if the first test after the $b_i$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.

@@

Therefore, this mechanism causes episode $i$ to be undetected if all of the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}. That is $b_i > \min(\sched_i)$ and $\tau_{\sched_i}(b_i) + b_i \leq e_i$.
    \item The episode ends in the interval $[\tau_{\sched_i}(b_i) + b_i, \tau^2_{\sched_i}(b_i) + b_i]$. Equivalently, $\tau_{\sched_i}(b_i) + 1 \leq d_i < \tau^2_{\sched_i}(b_i) + 1$.
    \item A false negative occurs on day $\tau_{\sched_i}(b_i) + b_i$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_i = b_i > \min(\sched_i)$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_i}(b_i) + 1 \leq D_i < \tau^2_{\sched_i}(b_i) + 1
    \mid b_i,
\theta \right) (1 - \psens) \\
&= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
\end{align}
Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as \cref{E-perf-test:eq:piu}, gives:
\begin{align}
\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
\end{align}

Denote the replacement for $p_{iu}$ as $p_{iu}'$.
Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$ is the sum of $p_{iu}$.
For the posterior density (\cref{E-perf-test:eq:full-posterior}), $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

@@

Here, I modify the simulation study in \cref{E-perf-test:sec:simulation-study} to assess several features of the statistical model I proposed in the previous section.
First, the ability to identify the survival function even with false negatives.
Second, the impact of misspecifying $\psens$, including violating the assumption that is constant.
Finally, the impact of the simplying approximations made in \cref{imperf-test:sec:modelling} for tractability.
These approximations are assuming that the previous negative is a true negative and that there is a negligible probability of missing an episode due to two false negatives.

I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0 (the latter being the same as the simulated data in \cref{E-perf-test:sec:simulation-study}), or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.

For each simulated dataset, I estimate the survival function using the model proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified.
However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
The amount that the simplifying assumptions are violated increases as $\psens$ decreases.

I use either the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
These were shown to be the best performing priors in \cref{E-perf-test:sec:results}).
Therefore, there are a total of 24 possible scenarios (each combination of: 4 simulated datasets, 3 values of $\psens$ in inference, and 2 priors).

When $\psens = 0.8$ and is correctly specified, the model recovers the true survival time well (see \cref{imperf-test:fig:constant-test-sensitivity}(B)).
The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplyifying assumptions, moving the estimated survival function closer to its true survival time.
However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}(A)).
This is likely caused by too large a violation of the simplifying assumptions made in \cref{imperf-test:sec:modelling}.
\begin{figure}
  % \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}}
  \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}
  \caption[Simulation study results with constant test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
    True survival time shown in black.
  }
  \label{imperf-test:fig:constant-test-sensitivity}
\end{figure}

If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates,
The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is underestimated.
This affect dominates over the opposing bias of overestimating the number of missed episodes.
The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
\begin{figure}
  % \makebox[\textwidth][c]{
    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-misspecified-sensitivity}
  \caption[Simulation study results with misspecified test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
    True survival time shown in black.
    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
    Hence (B) is correctly specified, and the same as in \cref{imperf-test:fig:constant-test-sensitivity}, but (A) and (C) have misspecified $\psens$.
    All results uses the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
  }
  \label{imperf-test:fig:misspecified-test-sensitivity}
\end{figure}

@@

In this section I apply the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
This is the 4,800 infection episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.

>>>>

Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$.
In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
If this is true, then the duration is $d_i = e_i - b_i + 1 = 1$.
Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
\todo{possibly a bit unclear phrasing here? struggling with it}
This produced the same pattern, that is estimating too many short episodes (not shown).

\Textcite{shenNonparametrica}, extending \textcite{panNote}, studied the situation when only the terminating event is interval censored and the initiating event has simple left truncation.
They showed that the maximum likelihood estimator is inconsistent if there is at least one single positive episode\todo{check the precise results and condition} and can lead to a severely biased underestimation of the survival function.
The CIS is a more complex setting, including double interval censoring and a complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
This means the result cannot simply be extended to the CIS, but still reinforces the intuition that single positive episodes are likely to be problematic.

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are two strands of evidence supporting this hypothesis.
First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Second, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see \cref{imperf-test:sec:simulate} for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this section, I introduce a probabilistic model for the generation of false negatives.
The simplest model for false negatives is a constant probability of returning a negative test even when detectable.
The probability of testing positive given being detectable is known as the \emph{test sensitivity}, $\psens$.
$\psens = 1$ means that there are no false negatives, the situation previously considered.
\todo{check if detectable and test sensitivity defined somewhere, and in a consistent way}
In \cref{E-perf-test}, the outcome of a test was deterministic: always positive if the individual being tested had an ongoing infection episode.
Now, the outcome of such a test result is a Bernoulli random variable with a probability $\psens$ of producing a positive.
If $\psens = 1$, this recovers the previous model without false negatives.
As previously, tests in individuals without ongoing infection episodes are always negative.
I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.

Following the modification, more single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
This is expected: a lower $\psens$ means that multiple positive tests are less likely, and hence the probability of a single positive episode is higher.

In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 93--96\%).
The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.

A better model would be to allow the rate of false negatives to vary over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

The CIS data provides evidence for a test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by classifying all test results as follows.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
  \label{imperf-test:eq:variable-test-sensitivity}
\end{equation}
where $t$ is the number of days since infection.
Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.

@@

First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability episode $i$ being admissible, to allow for the episode possibly being longer than observed.
Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability episode $i$ being undetected, to allow for additional episodes being missed.

@@

I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
If it is a false negative, then I will consider the episode's end right-censored.
However, if it is a true negative, then the episode's end is interval-censored, as previously.
A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.

I start by finding the tests that need consideration.
For tractability, assume that the negative test bounding the start of the infection, on day $l_i^{(b)}-1$, is a true negative.
This assumption is reasonable because the test sensitivity is high early in an infection; therefore, this test is unlikely to be a false negative.
For tractability and simplicity, I consider only tests between the positive tests providing a lower bound on the length of the episode.
By definition (see\todo{insert where these are defined in chapter 2}), these are the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive.
They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
Denote this subset as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq l_i^{(e)} \}$, and their results by $\vec{y_i'} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
As I assume that there are no false positives, the infection episode must span at least this period because it starts and ends with a positive test.
Therefore, the test results at times in $\sched_i'$ are either true positives or false negatives.

@@

First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
These occur with probability 1, by the assumption of no false positives.
The tests in $\sched_i'$ are either true positives or false negatives.
\begin{align}
&p(\vec{y_i'}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(\vec{y_i'}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(\vec{y_i'} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
\end{align}

@@

Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
I now consider the additional mechanisms arising due to false negatives.
Specifically, episode $i$ could be undetected if the first test after the $b_i$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.

@@

Therefore, this mechanism causes episode $i$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}. That is $b_i > \min(\sched_i)$ and $\tau_{\sched_i}(b_i) + b_i \leq e_i$.
    \item The episode ends in the interval $[\tau_{\sched_i}(b_i) + b_i, \tau^2_{\sched_i}(b_i) + b_i]$.
      Note that the lower bound here is exactly the bound on $e_i$ in the previous condition.
      Equivalently, $\tau_{\sched_i}(b_i) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_i) + 1$.
    \item A false negative occurs on day $\tau_{\sched_i}(b_i) + b_i$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_i = b_i > \min(\sched_i)$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_i}(b_i) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_i) + 1
    \mid b_i,
\theta \right) (1 - \psens) \\
&= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
\end{align}
Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as \cref{E-perf-test:eq:piu}, gives:
\begin{align}
\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
\end{align}

Denote the replacement for $p_{iu}$ as $p_{iu}'$.
Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$.
For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

@@

Here, I modify the simulation study in \cref{E-perf-test:sec:simulation-study} to assess several features of the statistical model I proposed in the previous section.
First, the ability to identify the survival function even with false negatives.
Second, the impact of misspecifying $\psens$, including violating the assumption that is constant.
Finally, the impact of the simplifying assumptions made in \cref{imperf-test:sec:modelling} for tractability.
These assumptions are that the previous negative is a true negative and that there is a negligible probability of missing an episode due to two false negatives.

I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0 (the latter being the same as the simulated data in \cref{E-perf-test:sec:simulation-study}), or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.

For each simulated dataset, I estimate the survival function using the model proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise I refer to it as misspecified.
However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
The amount that the simplifying assumptions are violated increases as $\psens$ decreases.

I use either the independent (vague) or model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
Therefore, there are a total of 24 possible scenarios (each combination of: 4 simulated datasets, 3 values of $\psens$ in inference, and 2 priors), although not all are shown.

When $\psens = 0.8$ and is correctly specified, the model recovers the true survival time well (see \cref{imperf-test:fig:constant-test-sensitivity}(B)).
The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplifying assumptions, moving the estimated survival function closer to its true survival time.
However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}(A)).
This is likely caused by too large a violation of the simplifying assumptions made in \cref{imperf-test:sec:modelling}.
\begin{figure}
  % \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}}
  \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}
  \caption[Simulation study results with constant test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
    True survival time shown in black.
  }
  \label{imperf-test:fig:constant-test-sensitivity}
\end{figure}

If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is underestimated.
This affect dominates over the opposing bias of overestimating the number of missed episodes.
The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
\begin{figure}
  % \makebox[\textwidth][c]{
    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-misspecified-sensitivity}
  \caption[Simulation study results with misspecified test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
    True survival time shown in black.
    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
    Hence, (B) is correctly specified, and the same as in \cref{imperf-test:fig:constant-test-sensitivity}, but (A) and (C) have misspecified $\psens$.
    All results use the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
  }
  \label{imperf-test:fig:misspecified-test-sensitivity}
\end{figure}

@@

In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
This is the 4,800 CIS infection episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
